---
title: "DEG_heatmap_optimized_&_Go_terms_K_means_EC"
output: html_document
date: "2025-06-13"
---

```{r}
library(Seurat)
library(dplyr)
library(pheatmap)
library(ggplot2)
library(tibble)

```
# K mean Heatmaps

Load your cell data

```{r}
EC <- readRDS("EC_cells.rds")
```

Create a 'condition' column

```{r}
EC@meta.data <- EC@meta.data %>%
  mutate(
    condition = case_when(
      origin == "A4T" ~ "4M_TAC",
      origin == "D4" ~ "4M",
      origin == "D28" ~ "28M",
      origin == "A24" ~ "24M", 
      origin == "Neo" ~ "Neo",
      TRUE ~ origin
    ),
    condition = factor(condition, levels = c("Neo", "4M", "4M_TAC", "24M", "28M"))
  )

# Differential expression
de_list <- list()
comparisons <- c("D28", "A4T", "Neo", "A24")

for (cond in comparisons) {
  de <- FindMarkers(EC, ident.1 = cond, ident.2 = "D4", group.by = "origin",
                    logfc.threshold = 0.25, min.pct = 0.1, only.pos = FALSE)
  de$gene <- rownames(de)
  de$comparison <- cond
  de_list[[cond]] <- de %>% filter(p_val_adj < 0.05)
}
```

Extract DEG gene sets

```{r}
all_degs <- bind_rows(de_list)
up_genes <- all_degs %>% filter(avg_log2FC > 0.25) %>% pull(gene) %>% unique()
down_genes <- all_degs %>% filter(avg_log2FC < -0.25) %>% pull(gene) %>% unique()
all_genes <- unique(c(up_genes, down_genes))
```

Subset and average expression

```{r}
EC_sub <- subset(EC, subset = origin %in% c("D4", "D28", "A4T", "Neo", "A24"))
avg_expr <- AverageExpression(EC_sub, group.by = "origin", return.seurat = FALSE)
avg_mat <- avg_expr$RNA[intersect(all_genes, rownames(avg_expr$RNA)), c("Neo", "A4T", "A24", "D28", "D4")]
avg_mat_scaled <- t(scale(t(avg_mat)))
```

Create Elbow plot

```{r}
wss <- sapply(1:15, function(k) {
  kmeans(avg_mat_scaled, centers = k, nstart = 25)$tot.withinss
})
```

K-means clustering on all DEGs (scaled)

```{r}
set.seed(123)
k <- 14
km <- kmeans(avg_mat_scaled, centers = k)
row_order <- order(km$cluster)
```

Plot Elbow plot

```{r}
#pdf("elbow_plot_all_DEGs.pdf")
plot(1:15, wss, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Plot for All DEGs")
#dev.off()
```

Plot DEG Expression heatmap (scaled Rows)

```{r}
#pdf("DEG_expression_heatmap_scaled_rows.pdf", width = 12, height = 8)
EC_all <- pheatmap(avg_mat_scaled[row_order, ],
         cluster_cols = TRUE, cluster_rows = FALSE,
         show_rownames = FALSE, show_colnames = TRUE,
         annotation_row = data.frame(Cluster = as.factor(km$cluster)),
         fontsize = 10, main = "DEG Expression Heatmap (Scaled Rows)")
#dev.off()
ggsave("DEG Expression Heatmap (Scaled Rows).pdf", plot = EC_all, width = 8, height = 6)
```

Convert to a data.frame with gene names and cluster assignment

```{r}
gene_clusters <- data.frame(
  gene = rownames(avg_mat_scaled),
  cluster = km$cluster
)

# Save to CSV
write.csv(gene_clusters, "gene_clusters_kmeans.csv", row.names = FALSE)
```

## Upregulated

```{r}
up_mat <- avg_mat[intersect(up_genes, rownames(avg_mat)), ]
up_mat_scaled <- t(scale(t(up_mat)))

wss_up <- sapply(1:10, function(k) {
  kmeans(up_mat_scaled, centers = k, nstart = 25)$tot.withinss
})

#pdf("elbow_plot_upregulated.pdf")
plot(1:10, wss_up, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "WSS",
     main = "Elbow Plot for Upregulated Genes")
#dev.off()


set.seed(123)
km_up <- kmeans(up_mat_scaled, centers = 5)
up_order <- order(km_up$cluster)

#pdf("upregulated_DEGs_heatmap_scaled_rows.pdf", width = 6, height = 8)
EC_up <-  pheatmap(up_mat_scaled[up_order, ],
         cluster_cols = TRUE, cluster_rows = FALSE,
         show_rownames = FALSE,
         annotation_row = data.frame(Cluster = as.factor(km_up$cluster)),
         fontsize = 10, main = "Upregulated DEGs (Scaled Rows)")
#dev.off()
ggsave("Upregulated DEGs_5 (Scaled Rows).pdf", plot = EC_up, width = 8, height = 6)

gene_clusters <- data.frame(
  gene = rownames(up_mat_scaled),
  cluster = km_up$cluster
)

write.csv(gene_clusters, "gene_clusters_up_kmeans_5.csv", row.names = FALSE)

```

## Downregulated

```{r}
down_mat <- avg_mat[intersect(down_genes, rownames(avg_mat)), ]
down_mat_scaled <- t(scale(t(down_mat)))

wss_down <- sapply(1:11, function(k) {
  kmeans(down_mat_scaled, centers = k, nstart = 25)$tot.withinss
})

#pdf("elbow_plot_downregulated.pdf")
plot(1:11, wss_down, type = "b", pch = 19,
     xlab = "Number of Clusters (k)",
     ylab = "WSS",
     main = "Elbow Plot for Downregulated Genes")
#dev.off()

set.seed(123)
km_down <- kmeans(down_mat_scaled, centers = 5)
down_order <- order(km_down$cluster)

#pdf("downregulated_DEGs_heatmap_scaled_rows.pdf", width = 6, height = 8)
EC_down <- pheatmap(down_mat_scaled[down_order, ],
         cluster_cols = TRUE, cluster_rows = FALSE,
         show_rownames = FALSE,
         annotation_row = data.frame(Cluster = as.factor(km_down$cluster)),
         fontsize = 10, main = "Downregulated DEGs (Scaled Rows)")
#dev.off()

ggsave("Downregulated DEGs_5 (Scaled Rows).pdf", plot = EC_down, width = 8, height = 6)

gene_clusters <- data.frame(
  gene = rownames(down_mat_scaled),
  cluster = km_down$cluster
)

write.csv(gene_clusters, "gene_clusters_down_kmeans_5.csv", row.names = FALSE)

```

# GO terms for K_means

Load necessary libraries

```{r}
library(clusterProfiler)
library(org.Mm.eg.db)
library(ggplot2)
```

## Upregulated:

Read the CSV file containing the upregulated DEGs and clusters

```{r}
gene_clusters <- read.csv("gene_clusters_up_kmeans_5.csv")

# View the first few rows to inspect
head(gene_clusters)

# Extract genes for each cluster
cluster1_genes <- gene_clusters$gene[gene_clusters$cluster == 1]
cluster2_genes <- gene_clusters$gene[gene_clusters$cluster == 2]
cluster3_genes <- gene_clusters$gene[gene_clusters$cluster == 3]
cluster4_genes <- gene_clusters$gene[gene_clusters$cluster == 4]
cluster5_genes <- gene_clusters$gene[gene_clusters$cluster == 5]


# Perform GO enrichment analysis
go_cluster1 <- enrichGO(gene = cluster1_genes, 
                        OrgDb = org.Mm.eg.db,   
                        keyType = "SYMBOL", 
                        ont = "BP",             
                        pvalueCutoff = 0.05)

go_cluster2 <- enrichGO(gene = cluster2_genes, 
                        OrgDb = org.Mm.eg.db,   
                        keyType = "SYMBOL", 
                        ont = "BP",             
                        pvalueCutoff = 0.05)

go_cluster3 <- enrichGO(gene = cluster3_genes, 
                        OrgDb = org.Mm.eg.db,
                        keyType = "SYMBOL", 
                        ont = "BP", 
                        pvalueCutoff = 0.05)

go_cluster4 <- enrichGO(gene = cluster4_genes, 
                        OrgDb = org.Mm.eg.db,
                        keyType = "SYMBOL", 
                        ont = "BP", 
                        pvalueCutoff = 0.05)

go_cluster5 <- enrichGO(gene = cluster5_genes, 
                        OrgDb = org.Mm.eg.db,
                        keyType = "SYMBOL", 
                        ont = "BP", 
                        pvalueCutoff = 0.05)

```

```{r, fig.width= 12, fig.height=8}
# Visualize enrichment results
barplot(go_cluster1, showCategory = 15) + ggtitle("GO Enrichment for Cluster 1")
#barplot(go_cluster2, showCategory = 15) + ggtitle("GO Enrichment for Cluster 2")
barplot(go_cluster3, showCategory = 15) + ggtitle("GO Enrichment for Cluster 3")
barplot(go_cluster4, showCategory = 15) + ggtitle("GO Enrichment for Cluster 4")
barplot(go_cluster5, showCategory = 15) + ggtitle("GO Enrichment for Cluster 5")
```

```{r}
# Convert enrichment results to data frames
df_cluster1 <- as.data.frame(go_cluster1)
df_cluster2 <- as.data.frame(go_cluster2)
df_cluster3 <- as.data.frame(go_cluster3)
df_cluster4 <- as.data.frame(go_cluster4)
df_cluster5 <- as.data.frame(go_cluster5)

# Preview geneID column (gene lists per GO term)
head(df_cluster1$geneID)

# Save each result to a CSV file
write.csv(df_cluster1, "GO_cluster1_results.csv", row.names = FALSE)
write.csv(df_cluster2, "GO_cluster2_results.csv", row.names = FALSE)
write.csv(df_cluster3, "GO_cluster3_results.csv", row.names = FALSE)
write.csv(df_cluster4, "GO_cluster4_results.csv", row.names = FALSE)
write.csv(df_cluster5, "GO_cluster5_results.csv", row.names = FALSE)

```

## Downregulated:

Read the CSV file containing the upregulated DEGs and clusters

```{r}
gene_clusters <- read.csv("gene_clusters_down_kmeans_5.csv")

# View the first few rows to inspect
head(gene_clusters)

# Extract genes for each cluster
cluster1_genes <- gene_clusters$gene[gene_clusters$cluster == 1]
cluster2_genes <- gene_clusters$gene[gene_clusters$cluster == 2]
cluster3_genes <- gene_clusters$gene[gene_clusters$cluster == 3]
cluster4_genes <- gene_clusters$gene[gene_clusters$cluster == 4]
cluster5_genes <- gene_clusters$gene[gene_clusters$cluster == 5]


# Perform GO enrichment analysis
go_cluster1 <- enrichGO(gene = cluster1_genes, 
                        OrgDb = org.Mm.eg.db,   
                        keyType = "SYMBOL", 
                        ont = "BP",             
                        pvalueCutoff = 0.05)

go_cluster2 <- enrichGO(gene = cluster2_genes, 
                        OrgDb = org.Mm.eg.db,   
                        keyType = "SYMBOL", 
                        ont = "BP",             
                        pvalueCutoff = 0.05)

go_cluster3 <- enrichGO(gene = cluster3_genes, 
                        OrgDb = org.Mm.eg.db,
                        keyType = "SYMBOL", 
                        ont = "BP", 
                        pvalueCutoff = 0.05)

go_cluster4 <- enrichGO(gene = cluster4_genes, 
                        OrgDb = org.Mm.eg.db,
                        keyType = "SYMBOL", 
                        ont = "BP", 
                        pvalueCutoff = 0.05)

go_cluster5 <- enrichGO(gene = cluster5_genes, 
                        OrgDb = org.Mm.eg.db,
                        keyType = "SYMBOL", 
                        ont = "BP", 
                        pvalueCutoff = 0.05)
```

```{r, fig.width= 12, fig.height=8}
# Visualize enrichment results
barplot(go_cluster1, showCategory = 15) + ggtitle("GO Enrichment for Cluster 1")
barplot(go_cluster2, showCategory = 15) + ggtitle("GO Enrichment for Cluster 2")
barplot(go_cluster3, showCategory = 15) + ggtitle("GO Enrichment for Cluster 3")
barplot(go_cluster4, showCategory = 15) + ggtitle("GO Enrichment for Cluster 4")
barplot(go_cluster5, showCategory = 15) + ggtitle("GO Enrichment for Cluster 5")

```

```{r} 
# Convert enrichment results to data frames 
df_cluster1 <-as.data.frame(go_cluster1) 
df_cluster2 <- as.data.frame(go_cluster2)
df_cluster3 <- as.data.frame(go_cluster3) 
df_cluster4 <-as.data.frame(go_cluster4) 
df_cluster5 <- as.data.frame(go_cluster5)

# Preview geneID column (gene lists per GO term)

head(df_cluster1$geneID)

# Save each result to a CSV file

write.csv(df_cluster1, "GO_cluster1_results.csv", row.names = FALSE)
write.csv(df_cluster2, "GO_cluster2_results.csv", row.names = FALSE)
write.csv(df_cluster3, "GO_cluster3_results.csv", row.names = FALSE)
write.csv(df_cluster4, "GO_cluster4_results.csv", row.names = FALSE)
write.csv(df_cluster5, "GO_cluster5_results.csv", row.names = FALSE)
