---
title: "Prepping data sets"
output: html_document
date: "2025-06-01"
---

```{r}
library(dplyr)
library(Seurat)
library(patchwork)
library(harmony)
library(ggplot2)
library(Matrix)
library(tibble)
library(readr)
library(cowplot)
library(clustree)
library(gridExtra)
library(stringr)
library(magrittr)
```
```{r}
#remotes::install_github('chris-mcginnis-ucsf/DoubletFinder', force = TRUE)
```
### Load hdf5 files from CellRanger for both D and L datasets
```{r}
inputdata.D <- Read10X_h5("D:/Doa_Multiome/filtered_feature_bc_matrix.h5")
inputdata.L <- Read10X_h5("D:/NeoandLPD/Data_files_neo_and_Adult/filtered_feature_bc_matrix.h5")
```

### Extract RNA data
```{r}
rna_counts.D <- inputdata.D$`Gene Expression`
rna_counts.L <- inputdata.L$`Gene Expression`
```
### Create Seurat objects
```{r}
D <- CreateSeuratObject(counts = rna_counts.D, project = "D")
L <- CreateSeuratObject(counts = rna_counts.L, project = "L")
```
### Add mitochondrial percentage
```{r}
D[["percent.mt"]] <- PercentageFeatureSet(D, pattern = "^mt-")
L[["percent.mt"]] <- PercentageFeatureSet(L, pattern = "^mt-")
```

#QC

### Perform QC filtering
```{r}
VlnPlot(D, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
VlnPlot(L, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
```

```{r}

D <- subset(D, subset = nFeature_RNA > 200 & nFeature_RNA < 4500 & percent.mt < 5)
L <- subset(L, subset = nFeature_RNA > 200 & nFeature_RNA < 5500 & percent.mt < 5)
VlnPlot(D, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
VlnPlot(L, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size = 0)
```


###Integrate objects
```{r}
Merged <- merge(D, y = L)
#or if you have a list: Merged <- merge(D, y = list(N, A, O))
```
```{r}
Merged$sample <- sub(".*-(\\d+)_.*", "\\1", rownames(Merged@meta.data))

Merged$origin <- paste(Merged$orig.ident, Merged$sample, sep = "_")

Merged@meta.data <- Merged@meta.data %>%
  mutate(origin = recode(origin, 
                         "D_1" = "D4",
                         "D_2" = "D4",
                         "D_3" = "D28",
                         "D_4" = "D28",
                         "L_1" = "Neo",
                         "L_2" = "A4T",
                         "L_3" = "A24"))
```




```{r}
sample_pal <- c("#006D2C", "#1D91C0", "#FEB24C", "#E31A1C",  "#800000")

names(sample_pal) <- c(
  "D4", "D28", "A4T", "Neo", "A24") 

pal <- c("#66C2A4", "#006D2C", "#1D91C0","#225EA8","#081D58","#f79736", "#E31A1C", "#800000" )

#names(dcell_pal) <- c(
#  "CM", "EC", "FB", "MACRO", "PERI", "IC", "SMC") 

# Plot Percent of mitochondrial genes by Sample
VlnPlot(Merged,
        group.by = "origin",
        features = "percent.mt",
        # log = TRUE,
        pt.size = 0, 
) +
  NoLegend() +
  scale_fill_manual(values = sample_pal)

#VlnPlot(Merged,group.by = "origin",features = "perc.ribo",
        # log = TRUE,
#        pt.size = 0, 
#) +
#  NoLegend() +
#  scale_fill_manual(values = sample_pal)

#VlnPlot(Merged,
#        group.by = "origin",
#        features = "perc.hb",
#        # log = TRUE,
#        pt.size = 0, 
#) +
#  NoLegend() +
#  scale_fill_manual(values = sample_pal)

# Plot features by Sample together
#VlnPlot(Merged,
#        group.by = "origin",
#        features = c("percent.mt", "perc.ribo", "perc.hb"),
#        # log = TRUE,
#        pt.size = 0
#) &
#  NoLegend() &
#  scale_fill_manual(values = sample_pal)

# Comparison of mitochondrial and hemoglobin genes
#p1 <- Merged@meta.data %>%
#  ggplot(., aes(x = percent.mt, y = perc.hb, color = celltype)) +
#  geom_point() +
#  scale_color_manual(values = pal) +
#  theme_classic()
#p2 <- Merged@meta.data %>%
#  ggplot(., aes(x = percent.mt, y = perc.hb, color = `origin`)) +
#  geom_point() +
#  scale_color_manual(values = sample_pal) +
#  theme_classic() 

#p1 | p2



# Percent Mitochondrial Genes
Merged@meta.data %>%
  ggplot(., aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) +
  geom_point() +
  theme_classic() +
  scale_color_gradient(low = "yellow", high = "red")

# Percent MT Genes per sample
Merged@meta.data %>%
  ggplot(., aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) +
  geom_point() +
  facet_wrap(~`origin`) +
  theme_classic() +
  scale_color_gradient(low = "yellow", high = "red") +
  scale_x_continuous(
    labels = scales::unit_format(unit = "K", scale = 1e-3))

# Plot by cell type
Merged@meta.data %>%
  ggplot(., aes(x = nCount_RNA, y = nFeature_RNA, color = origin)) +
  geom_point() +
  theme_classic() + 
  scale_color_manual(values = pal)

```
##RAN Scrublet in Python


#Merged <- Merged %>%
#  NormalizeData(verbose = FALSE) %>%
#  FindVariableFeatures(method = "vst", nfeatures = 3000, verbose = FALSE) %>%
#  ScaleData(verbose = FALSE, features = VariableFeatures(.)) %>%
#  RunPCA(verbose = FALSE) %>%
#  RunHarmony(group.by.vars = "orig.ident", plot_convergence = TRUE) %>%  # Apply Harmony
#  FindNeighbors(dims = 1:10, reduction = "harmony") %>%  # Use Harmony reduction
#  FindClusters(resolution = 0.5) %>%
#  RunUMAP(dims = 1:10, reduction = "harmony")  # Use Harmony reduction for UMAP

```{r} 
Merged <- readRDS("D:/NeoandLPD/Multiome/Singlet/Merged_Final.rds")
Merged <- NormalizeData(Merged)
Merged <- FindVariableFeatures(Merged, selection.method = "vst", nfeatures = 2000)
all.genes <- rownames(Merged)
Merged <- ScaleData(Merged, features = VariableFeatures(object = Merged))
Merged <- RunPCA (Merged, features=VariableFeatures(object = Merged))
ElbowPlot(Merged)
Merged <- RunHarmony(Merged,group.by.vars = "orig.ident", plot_convergence = TRUE)
Merged <- FindNeighbors(Merged, dims = 1:20, reduction = "harmony")  # Use Harmony reduction
Merged <- FindClusters(Merged, resolution = 0.5) 
Merged <- RunUMAP(Merged, dims = 1:10, reduction = "harmony")  # Use Harmony reduction for UMAP

#saveRDS(Merged, file = "Merged_arc.rds")

```

## Annotations
  
###individual clusters
```{r, fig.width=8, fig.height=6}
Idents(Merged) <- "seurat_clusters"
DimPlot(Merged, reduction = "umap", label = TRUE)
```
```{r, fig.width=14, fig.height=6}
marker_genes <- list(
  CM = c("Tnnt2", "Actc1", "Myh6", "Myh7", "Myl2", "Myl7"),
  EC = c("Pecam1", "Cdh5", "Vegfa", "Kdr", "Nos3"),
  SMC = c("Acta2", "Tagln", "Myh11", "Smtn"),
  FB = c("Col1a1", "Col3a1", "Postn", "Dcn", "Tcf21"),
  Macro = c("Adgre1", "Cd68", "Itgam", "Csf1r", "Mrc1"),
  B_cells = c("Cd19", "Cd79a", "Ms4a1"),   
  T_cells = c("Cd3e", "Cd4", "Cd8a", "Il7r"),
  Meso = c("Msln", "Wt1"),
  Pericytes = c("Pdgfrb", "Cspg4", "Anpep"),
  Mono = c("Cd14", "Ms4a7"),
  NK = c("Nkg7"),
  DC = c("Cst3")
)
DotPlot(Merged, features = marker_genes) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
Idents(Merged) <- "seurat_clusters"

new.cluster.ids <- c("CM", "EC", "FB", "CM", "MACRO", "CM", "PERI", "EC", "EC", "B", "EC", "T", "CM", "MACRO", "CM", "SMC")
names(new.cluster.ids) <- levels(Merged)
Merged <- RenameIdents(Merged, new.cluster.ids)
DimPlot(Merged, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()

Idents(Merged) <- new.cluster.ids
```




```{r, fig.width=14, fig.height=6}
# Plot violin plots for marker genes
Idents(Merged) <- "seurat_clusters"
for (celltype in names(marker_genes)) {
  plot <- VlnPlot(Merged, features = marker_genes[[celltype]], pt.size = 0.0, ncol = 2)
  print(plot)  # show plot in the output
  # Optional: ggsave(paste0(celltype, "_violin_plot.png"), plot = plot, width = 20, height = 10)
}
```

```{r, fig.width=6, fig.height=6}
# Calculate module scores for each cell type
for (cell_type in names(marker_genes)) {
  Merged <- AddModuleScore(Merged, features = list(marker_genes[[cell_type]]), name = cell_type)
}

# Loop through each module and plot UMAP
for (cell_type in names(marker_genes)) {
  p <- FeaturePlot(Merged, features = paste0(cell_type, "1"), min.cutoff = "q10", max.cutoff = "q90") +
    ggtitle(cell_type)
  print(p)  # show plot in the output
  # Optional: ggsave(filename = paste0(cell_type, "_UMAP.png"), plot = p, width = 6, height = 5, dpi = 300)
}
```

```{r, fig.width=16, fig.height=15}
vln_plot <- VlnPlot(
  Merged,
  features = paste0(names(marker_genes), "1"),
  group.by = "seurat_clusters",
  pt.size = 0,
  ncol = 2
) + theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(vln_plot)
# Optional: ggsave("ModuleScores_ViolinPlot.png", plot = vln_plot, width = 12, height = 6, dpi = 300)


```
